---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# DAGassist

<!-- badges: start -->
<!-- badges: end -->

**An all-in-one DAG-driven robustness check.**
Classifies variables by causal role, computes the smallest back-door adjustment set, and compares the significance of the original and minimal models.


## Installation

You can install the development version of DAGassist from [GitHub](https://github.com/) with:

``` {r install, eval = FALSE}
install.packages("pak")
pak::pak("grahamgoff/DAGassist")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(DAGassist) 

dag_assist(dag = test_complex, #this example DAG ships with the package
           formula = Y ~ X + M + C + Z,
           engine = lm,
           data = test_df, # this example dataset ships with the package
           exposure = "X",
           outcome = "Y")
```

