---
title: "Total and direct effects"
bibliography: ../references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150
)
options(DAGassist.color = FALSE, crayon.enabled = FALSE)
```

Regression estimates a statistical quantity of interest called an **estimand**, which is implied by the model's controls. As the [homepage](../index.html) shows, controlling for a mediator transforms a total effect into something close to a direct effect. The `estimand` argument allows users to explicitly target an estimand of interest [@LundbergJohnsonStewart2021]: `"total"` targets the average total effect, and `"direct"` targets the average controlled direct effect. This article explains each column `DAGassist` adds, how it is estimated, and what it assumes.

```{r setup}
library(DAGassist)
```

The turnout example supports two different questions:

- **Total effect.** How much would turnout change if income rose by one unit, through every path, including by raising political interest? In `turnout_data`, the answer is 0.50.
- **Controlled direct effect.** How much would turnout change if income rose by one unit *while political interest were held fixed*? The answer is 0.30.

Requesting both estimands adds a set of columns (minimal and canonical) for each:

```{r both-estimands, results = "asis"}
DAGassist(turnout_dag,
          lm(turnout ~ income + state + age + polint + industry + elect_comp,
             data = turnout_data),
          estimand = c("total", "direct"),
          show = "models", type = "text", verbose = FALSE)
```

| Column | Question it answers | Estimate | Truth |
|:--|:--|:-:|:-:|
| Original | Whatever your controls imply | 0.28 | |
| Total … (Raw) | Total effect, by regression adjustment | 0.49 | 0.50 |
| Total … (Weighted) | Total effect, by inverse-probability weighting | 0.49–0.50 | 0.50 |
| Direct (Raw) | Controlled direct effect, by sequential g-estimation | 0.28 | 0.30 |
| Direct (Weighted) | The same, with inverse-probability weights | 0.28 | 0.30 |

The original regression's 0.28 is close to the direct effect, because it controls for the mediator `polint`. The weighted total-effect columns report only the exposure's effect, so the other rows are blank.

## How the total-effect columns are estimated

The **raw** columns fit your model with each DAG-derived adjustment set (see [Causal roles and adjustment sets](roles-and-sets.html)). In a linear model with no interactions involving the exposure, the exposure's coefficient is the average total effect.

The **weighted** columns take three steps:

1. Model the exposure given that set's controls, using [WeightIt](https://ngreifer.github.io/WeightIt/). For a binary exposure this is a propensity score; for a continuous one, a generalized propensity score.
2. Refit the outcome model with the resulting inverse-probability weights.
3. Report the exposure's average marginal effect, using [marginaleffects](https://marginaleffects.com/).

Regression adjustment relies on a correctly specified outcome model; weighting relies on a correctly specified exposure model. When the two agree, as they do here, neither kind of misspecification is driving the result. When they disagree, at least one of the models is wrong.

## Weight diagnostics

With `estimand = "total"`, the console report ends with a block of weight diagnostics:

```{r weight-diagnostics, echo = FALSE}
report <- DAGassist(turnout_dag,
                    lm(turnout ~ income + state + age + polint + industry + elect_comp,
                       data = turnout_data),
                    estimand = "total")
out <- capture.output(print(report))
start <- grep("^Weight diagnostics", out)
end <- which(out == "")
end <- end[end > start][1] - 1
cat(out[start:end], sep = "\n")
```

`w range` is the smallest and largest weight. `ESS` is the Kish effective sample size, $(\sum w)^2 / \sum w^2$: the number of equally weighted observations the weighted sample is worth. `DAGassist` flags two problems:

- `LOW_ESS`: the effective sample size is less than half the actual sample.
- `EXTREME_W`: the largest weight is more than 20 times the median weight.

Both flags fire here. The 5,000 observations are worth about 1,400 once weighted, and a handful of observations carry weights near 70. This is common with continuous exposures, whose density-based weights are notoriously variable [@NaimiEtAl2014]. The weighted estimates are still close to the truth, but they are less precise than the raw ones.

<!-- NOTE: the sentence above requires the weighted-SE fix in R/weights.R (vcov = "HC0"); until then the table shows weighted SEs that are too small. -->

Two remedies are available. Setting `weights_args = list(trim_at = 0.99)` caps the weights at their 99th percentile, trading a little bias for a large reduction in variance. Alternatively, when the outcome model is credible, give more weight to the raw columns.

## How the direct-effect columns are estimated

The **controlled direct effect** is the effect of the exposure on the outcome if every unit's mediators were set to the same value.

In the turnout example, you could estimate it by simply controlling for the mediator. That works because nothing income affects also confounds political interest and turnout. When such a variable exists, an *intermediate confounder*, every ordinary regression is biased. Controlling for it blocks part of the direct effect; leaving it out opens a non-causal path through the mediator.

Sequential g-estimation avoids both problems in two stages [@AcharyaBlackwellSen2016]. First, it estimates the mediator's effect on the outcome, controlling for intermediate confounders. Second, it subtracts that effect from the outcome and regresses what remains on the exposure and the baseline covariates only.

`DAGassist` builds the call to `DirectEffects::sequential_g()` from your DAG and formula, in the form `outcome ~ exposure + baseline | intermediate confounders | mediators`:

- Mediator terms are centered, and non-binary mediators also get a squared term, so the first stage isn't forced to be linear.
- **Direct (Weighted)** adds inverse-probability weights estimated from the baseline covariates only, never from the mediators.
- Standard errors come from a two-stage sandwich estimator, clustered if you supply a cluster variable.

### An example with an intermediate confounder

In the simulated data below, the exposure `A` affects `L`, and `L` affects both the mediator `M` and the outcome `Y`.

```{r sim-data}
set.seed(1)
n <- 5000
C <- rnorm(n)                                          # baseline confounder
A <- 0.5 * C + rnorm(n)                                # exposure
L <- 0.6 * A + 0.4 * C + rnorm(n)                      # intermediate confounder
M <- 0.5 * A + 0.5 * L + rnorm(n)                      # mediator
Y <- 0.3 * A + 0.4 * M + 0.5 * L + 0.3 * C + rnorm(n)  # outcome
sim <- data.frame(A, C, L, M, Y)

sim_dag <- ggdag::dagify(
  Y ~ A + M + L + C,
  M ~ A + L,
  L ~ A + C,
  A ~ C,
  exposure = "A",
  outcome  = "Y"
)
```

```{r sim-dag-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 3.2, fig.alt = "DAG: C causes A, L, and Y. A causes L, M, and Y. L causes M and Y. M causes Y."}
library(ggdag)
library(ggplot2)

dagitty::coordinates(sim_dag) <- list(
  x = c(A = 0, L = 2, M = 4, Y = 6, C = 3),
  y = c(A = 0, L = 1.3, M = 1.3, Y = 0, C = -1.2)
)

#coordinates for dag labels
label_df <- data.frame(
  name = c("A", "L", "M", "Y", "C"),
  label_x = c(-0.9, 2, 4, 6.8, 3),
  label_y = c(0, 1.75, 1.75, 0, -1.2),
  label = c("A (exposure)", "L (intermediate confounder)", "M (mediator)",
            "Y (outcome)", "C (baseline confounder)")
)

ggdag_status(sim_dag, text = FALSE) +
  theme_dag() +
  scale_color_manual(values = c(exposure = "black", outcome = "grey30"),
                     na.value = "grey60") +
  theme(legend.position = "none") +
  geom_label(
    data = label_df,
    aes(x = label_x, y = label_y, label = label),
    inherit.aes = FALSE,
    size = 3.4,
    label.size = 0.25
  )
```

Holding `M` fixed, `A` still affects `Y` directly (0.3) and through `L` (0.6 × 0.5 = 0.3), so the controlled direct effect with respect to `M` is **0.60**. No ordinary regression recovers it:

```{r sim-regressions}
coef(lm(Y ~ A + M + C, data = sim))[["A"]]      # leaves out L
coef(lm(Y ~ A + M + L + C, data = sim))[["A"]]  # controls for L
```

Leaving out `L` conditions on `M`, which opens the non-causal path `A -> M <- L -> Y`. Controlling for `L` blocks `A -> L -> Y`, which is part of the direct effect.

By default, `DAGassist` holds *every* mediator in your formula fixed. `L` is itself a mediator (`A -> L -> Y`), so the default estimates the effect of `A` through neither `L` nor `M`, which is 0.30:

```{r sim-default, results = "asis"}
DAGassist(sim_dag, lm(Y ~ A + M + L + C, data = sim),
          estimand = "direct", show = "models", type = "text", verbose = FALSE)
```

To hold only `M` fixed, tell `DAGassist` which variable is the mediator and which is the intermediate confounder with the `acde` argument:

```{r sim-override, results = "asis"}
DAGassist(sim_dag, lm(Y ~ A + M + L + C, data = sim),
          estimand = "direct", acde = list(m = "M", z = "L"),
          show = "models", type = "text", verbose = FALSE)
```

Sequential g-estimation now recovers the controlled direct effect of 0.60. `acde` also accepts `x` (baseline covariates) and `fe` (fixed effects) for full control over the specification.

## Assumptions

Both estimands assume the DAG is right. They also rest on assumptions that no model comparison can check:

- **No unmeasured confounding of the exposure and outcome.** See [Robustness to DAG uncertainty](robustness.html).
- **For the direct effect, no unmeasured confounding of the mediator and outcome either.** The robustness checks examine the total-effect adjustment sets only, so they won't flag this.
- **No exposure–mediator interaction.** `DAGassist`'s first stage has no interaction term, so it assumes the direct effect is the same at every value of the mediator. If you expect an interaction, fit `DirectEffects::sequential_g()` directly.
- **Positivity.** Every unit must have some chance of receiving each exposure level, given its covariates. The weight diagnostics are the practical check: extreme weights mean some units have almost no comparable counterparts.

Estimand recovery currently supports a single exposure, which cannot be an interaction term. To study an interaction, compute it as its own variable first.

## Which estimand to report

- Use `estimand = "total"` when you are interested in what happens if the exposure changes, irrespective of mechanisms.
- Use `estimand = "direct"` when you are interested in how much of the effect is unmediated.

## References
