---
title: "Robustness to DAG uncertainty"
bibliography: ../references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150
)
options(DAGassist.color = FALSE, crayon.enabled = FALSE)
```

Every DAG makes two kinds of assumptions that are difficult to settle empirically: 1) the direction of each edge, and 2) the absence of every edge that isn't drawn. A missing edge implies that one variable has no direct effect on another [@HaberEtAl2022]. `DAGassist` checks whether your adjustment sets, and therefore your estimate, depend on those assumptions. This article continues the voter-turnout example from the [homepage](../index.html) and [Causal roles and adjustment sets](roles-and-sets.html).

```{r setup}
library(DAGassist)
```

## Uncertain arrow directions

The turnout DAG assumes that income raises political interest (`income -> polint`), which makes political interest a mediator. However, it is also plausible that politically engaged people may seek out better-paid work. The dashed line marks the arrow whose direction is in doubt.

```{r uncertain-dag, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 3.5, fig.alt = "The turnout DAG. Age and parental_ses affect income and turnout; industry affects income; election competitiveness affects turnout; political interest affects turnout. The link between income and political interest is drawn as a dashed line with no arrowhead because its direction is uncertain."}
library(ggdag)
library(ggplot2)

# plot the DAG without the uncertain edge, then draw that edge as a dashed line
plot_dag <- dagitty::dagitty(sub("income -> polint\n", "",
                                 as.character(turnout_dag), fixed = TRUE))

#coordinates for dag labels
label_df <- data.frame(
  name = c("turnout", "income", "parental_ses", "age", "polint", "industry", "elect_comp"),
  label_x = c(10, 0, 5, 5, 5, 0, 9),
  label_y = c(0, 0, 1, -1, 0.55, -0.75, -0.75)
)
label_df$label <- unname(attr(turnout_dag, "labels")[label_df$name])

ggdag_status(plot_dag, text = FALSE) +
  theme_dag() +
  scale_color_manual(values = c(exposure = "black", outcome = "grey30"),
                     na.value = "grey60") +
  theme(legend.position = "none") +
  annotate("segment", x = 0.5, y = 0.05, xend = 4.5, yend = 0.45,
           linetype = "dashed", linewidth = 0.9, colour = "#AC4848") +
  annotate("label", x = 2.5, y = 0.25, label = "?", colour = "#AC4848",
           fontface = "bold", size = 4, label.size = 0) +
  geom_label(
    data = label_df,
    aes(x = label_x, y = label_y, label = label),
    inherit.aes = FALSE,
    size = 3.4,
    label.size = 0.25
  )
```

`pdag_robustness()` takes the DAG and the edges whose direction you're unsure of, written `A -- B`. It considers every acyclic orientation of those edges and then reports whether the adjustment sets, or any variable's role, change across orientations.

```{r pdag-one}
f <- turnout ~ income + parental_ses + age + polint + industry + elect_comp

pdag_robustness(turnout_dag, formula = f, uncertain_edges = "income -- polint")
```

Both adjustment sets change, and `polint` flips between mediator (a bad control) and confounder (a good one). The formula is optional. When you supply it, `DAGassist` checks whether the variables whose roles change are actually in your model, which determines whether it recommends re-estimating.

You can list several uncertain edges at once:

```{r pdag-two}
pdag_robustness(turnout_dag, formula = f,
                uncertain_edges = c("income -- polint"))
```

Each uncertain edge doubles the number of possible orientations, so `max_uncertain` caps the list at 10 edges (1,024 worlds) by default. Cyclical orientations are skipped because they are not valid DAGs.

## Re-estimation

`DAGassist` recommends re-estimating when, as above, edge uncertainty affects adjustment sets. To see the estimate under the alternative orientation, reverse the arrow and rerun `DAGassist()`:

```{r alt-models, results = "asis"}
alt_dag <- dagitty::dagitty(sub("income -> polint", "polint -> income",
                                as.character(turnout_dag), fixed = TRUE))

DAGassist(alt_dag,
          lm(turnout ~ income + parental_ses + age + polint + industry + elect_comp,
             data = turnout_data),
          show = "models", type = "text", verbose = FALSE)
```

If political interest causes income, `polint` is a confounder, the minimal set becomes `{age, polint, parental_ses}`, which yields an estimate of about 0.28. Under the original DAG, `polint` is a mediator and the model returns an estimate of 0.49. The data are identical in both cases; the answer depends entirely on the direction of the edge between `income` and `polint`.

## Missing arrows

`add_edges_robustness()` asks the complementary question: what if the DAG omits out an edge that exists in the underlying data generating process? It adds each candidate edge to the DAG one at a time and reports how the adjustment sets and roles change. A directed edge (`A -> B`) is an omitted direct effect; a bidirected edge (`A <-> B`) is an unmeasured common cause of `A` and `B`.

```{r add-edges}
add_edges_robustness(turnout_dag, formula = f,
                     add_edges = c("industry -> turnout",
                                   "polint <-> turnout",
                                   "income <-> turnout"))
```

- **`industry -> turnout`.** If industry affects turnout directly, it becomes a confounder, and the minimal set must include it. The canonical set doesn't change, because it already contains `industry`. Estimates from the canonical model are therefore robust to this particular omission.
- **`polint <-> turnout`.** An unmeasured common cause of political interest and turnout leaves the total-effect adjustment sets unchanged.
- **`income <-> turnout`.** An unmeasured common cause of the exposure and the outcome cannot be removed by any set of controls. If that confounding is plausible, adjustment alone can't recover the effect.

## Running robustness checks inside `DAGassist()`

Edge orientation and missing edge diagnostics are available as arguments to `DAGassist()`, which appends them to the standard report. `uncertain_edges` and `add_edges` work as above, and `pdag` accepts a `dagitty` PDAG whose undirected (`--`) edges are treated as uncertain.

```{r in-report}
DAGassist(turnout_dag,
          lm(turnout ~ income + parental_ses + age + polint + industry + elect_comp,
             data = turnout_data),
          show = "roles",
          uncertain_edges = "income -- polint",
          add_edges = "industry -> turnout",
          verbose = FALSE)
```

The standalone functions only require a DAG and a formula, so you can run them before collecting any data. `DAGassist()` also needs the data, because it fits the comparison models.

## Choosing which edges to test

Testing every possible edge is neither feasible nor informative. Good candidates are:

- **Arrows whose timing is ambiguous**, where reverse causation is plausible (income and political interest; income and place of residence).
- **Unmeasured common causes** of the exposure and the outcome.

## References
