---
title: "Causal roles and adjustment sets"
bibliography: ../references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150
)
options(DAGassist.color = FALSE, crayon.enabled = FALSE)
```

```{r,echo=FALSE}
devtools::load_all(quiet = TRUE)
```


Whether a control variable belongs in a regression depends on its relationships with other variables. `DAGassist` labels each variable with a causal role, then builds the adjustment sets those roles imply. This article explains every role, which ones to adjust for, and how `DAGassist` builds the minimal and canonical adjustment sets. Our causal role classification framework follows @CinelliForneyPearl2022's taxonomy of good, bad, and neutral controls.

```{r setup, eval=FALSE}
library(DAGassist)
```

## A DAG with every role

To see every role at once, we write a small DAG in `dagitty` syntax. Each letter stands for one role: `X` is the exposure and `Y` the outcome.

```{r roles-dag}
roles_dag <- dagitty::dagitty("dag {
  X [exposure]
  Y [outcome]
  X -> Y
  Z -> X
  Z -> Y
  Z -> V -> Y
  Z -> W
  X -> M -> Y
  M -> N
  X -> C <- Y
  C -> Q
  Y -> D
  E -> X
  F -> Y
}")
```

```{r roles-dag-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4, fig.alt = "DAG with exposure X and outcome Y. Z (confounder) causes X and Y, and causes V, which causes Y, and W. X causes M (mediator), which causes Y and N. X and Y both cause C (collider), which causes Q. Y causes D. E causes X only; F causes Y only."}

library(dagitty)
library(ggdag)
library(ggplot2)

x_pos <- c(
  X = 0,
  Y = 10,
  Z = 5,
  V = 8,
  W = 2,
  E = 0,
  F = 10,
  M = 3.5,
  N = 3.5,
  C = 6.5,
  Q = 6.5,
  D = 10
)

y_pos <- c(
  X = 0,
  Y = 0,
  Z = 2,
  V = 1.5,
  W = 2.6,
  E = 1.5,
  F = 1.5,
  M = -1.2,
  N = -2.2,
  C = -1.2,
  Q = -2.2,
  D = -1.5
)


roles_dag <- dagify(
  
  Y ~ X + Z + V + M + F,
  
  X ~ Z + E,
  
  V ~ Z,
  
  W ~ Z,
  
  M ~ X,
  
  N ~ M,
  
  C ~ X + Y,
  
  Q ~ C,
  
  D ~ Y,
  
  exposure = "X",
  outcome  = "Y",
  
  coords = list(x = x_pos, y = y_pos),

  labels = c(
    X = "X (exposure)",
    Y = "Y (outcome)",
    Z = "Z (conf)",
    V = "V (dConfOn)",
    W = "W (dConfOff)",
    E = "E (NCT)",
    F = "F (NCO)",
    M = "M (med)",
    N = "N (dMed)",
    C = "C (col)",
    Q = "Q (dCol)",
    D = "D (dOut)"
  )
)


#coordinates for dag labels
label_df <- data.frame(
  name = c("X", "Y", "Z", "V", "W", "E", "F", "M", "N", "C", "Q", "D"),
  label_x = c(-1.4, 11.4, 5, 8, 2, 0, 10, 3.5, 3.5, 6.5, 6.5, 10),
  label_y = c(0, 0, 2, 1.95, 2.6, 1.5, 1.5, -1.2, -2.2, -1.2, -2.2, -1.5)
)

# pull labels safely from dagitty object
labels_vec <- attr(roles_dag, "labels")

label_df$label <- unname(labels_vec[label_df$name])

# ---- plot: ggdag without labels, then add manual labels
p <- ggdag_status(roles_dag, text = FALSE) +
  theme_dag() +
  scale_fill_manual(values = c(exposure="black", outcome="grey30", none="grey90")) +
  scale_color_manual(values = c(exposure="black", outcome="grey30", none="grey60")) +
  theme(legend.position = "none") +
  geom_label(
    data = label_df,
    aes(x = label_x, y = label_y, label = label),
    inherit.aes = FALSE,   
    size = 3.4,
    label.size = 0.25
  )

p
```


## Reading the roles table

Roles depend only on the DAG, so this step needs no data:

```{r roles-table}
DAGassist(roles_dag, show = "roles", verbose = FALSE)
```

Each column is a role, and an `x` marks the roles a variable plays. The table below defines each role and the causal inference implications of adjusting for each when the target is the total effect of `X` on `Y`. Note that a variable can play more than one role, but the `role` column reports the most identification-relevant one. For example, if a variable were a descendant of a mediator and also a neutral control on the outcome, its primary role would be mediator because mediators shift estimands and neutral controls do not.

| Column | Role | Example | What it is | Adjust for it? |
|:--|:--|:-:|:--|:--|
| `conf` | Confounder | `Z` | A common cause of `X` and `Y` | **Yes.** It opens a back-door path (`X <- Z -> Y`) that must be blocked. |
| `dConfOn` | Descendant of a confounder, on a back-door path | `V` | Caused by a confounder, and causes `Y` | **Safe.** It blocks the part of the back-door path that runs through it. |
| `NCO` | Neutral control on the outcome | `F` | Causes `Y` but not `X` | **Safe**, and usually improves precision. |
| `NCT` | Neutral control on the treatment | `E` | Causes `X` but not `Y` | **Safe for bias**, but usually widens standard errors. |
| `dConfOff` | Descendant of a confounder, off the back-door path | `W` | Caused by a confounder, but doesn't affect `Y` | **Irrelevant.** It neither removes nor adds bias. |
| `med` | Mediator | `M` | On a causal path, `X -> M -> Y` | **No.** It removes part of the effect, turning the total effect into a direct effect. |
| `dMed` | Descendant of a mediator | `N` | Caused by a mediator | **No.** It partly controls for the mediator. |
| `col` | Collider | `C` | Caused by both `X` and `Y` | **No.** It opens a non-causal path between `X` and `Y`. |
| `dCol` | Descendant of a collider | `Q` | Caused by a collider | **No.** It partly opens the same path. |
| `dOut` | Descendant of the outcome | `D` | Caused by `Y` | **No.** It biases the estimate. |


## Minimal and canonical adjustment sets

An adjustment set is a set of controls that blocks every back-door path from `X` to `Y` without including anything `X` affects. `DAGassist` finds these sets with `dagitty` [@TextorEtAl2016] and fits one model for each:

- A **minimal** set contains no redundant variables: remove any one, and some back-door path reopens. A DAG can have several minimal sets, and `DAGassist` fits a model for each ("Minimal 1", "Minimal 2", and so on).
- The **canonical** set contains every variable that causes `X` or `Y`, except those the exposure affects [@vanderZanderEtAl2014]. It is unique, and it is a valid adjustment set whenever any valid set exists.

For `roles_dag`:

```{r roles-sets}
dagitty::adjustmentSets(roles_dag, type = "minimal")
dagitty::adjustmentSets(roles_dag, type = "canonical")
```

The minimal set is just the confounder. The canonical set adds `V` and the two
neutral controls, and leaves out everything marked "do not adjust" as well as
the irrelevant `W`.

| Role | Example | Minimal | Canonical |
|:--|:-:|:-:|:-:|
| Confounder | `Z` | ✓ | ✓ |
| Descendant of a confounder, on a back-door path | `V` | \* | ✓ |
| Neutral control on the outcome | `F` | | ✓ |
| Neutral control on the treatment | `E` | | ✓ |
| Descendant of a confounder, off the back-door path | `W` | | |
| Mediator, descendant of a mediator | `M`, `N` | ✗ | ✗ |
| Collider, descendant of a collider | `C`, `Q` | ✗ | ✗ |
| Descendant of the outcome | `D` | ✗ | ✗ |

✓ = included. Blank = left out, but harmless to add. ✗ = left out because
adjusting for it biases the total effect.

\* `V` could replace `Z` in a minimal set only if every back-door path through
`Z` also ran through `V`. In `roles_dag`, `Z` also affects `Y` directly, so it
can't.

## Why the two sets can give different answers

Both sets remove confounding, so both should recover the same effect. They
differ in precision. The turnout example on the [homepage](../index.html)
shows this: its canonical set adds a neutral control on the outcome
(`elect_comp`) and a neutral control on the treatment (`industry`) to the
minimal set `{age, state}`.

```{r turnout-models, results = "asis"}
DAGassist(turnout_dag,
          lm(turnout ~ income + state + age + polint + industry + elect_comp,
             data = turnout_data),
          show = "models", type = "text", verbose = FALSE)
```

The minimal and canonical estimates agree (0.49), and both differ sharply from
the original model, which controls for the mediator `polint`. Here the
canonical standard error is slightly smaller (0.015 vs. 0.016). If the DAG is
right, the minimal and canonical estimates should differ only by sampling
noise, so a large gap between them suggests the DAG is wrong somewhere. The
[robustness article](robustness.html) shows how to probe that.

## References
